# Lab 1C; Lists and dataframes {#lab1crecursivevectors}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
```{r}
library(tidyverse)
```

## Atomic vectors vs. recursive vectors
The R vectors above are so called `atomic vectors`. They are the 'atoms' of R, the building block of which all other objects can be build. The resulting constituated R objects that can be build from the 'atoms' are called `recursive vectors`. The molecules. With these atoms and molecules we can build really complex data objects.

## Recursive vectors
The recursive vectors are:

 * List
 * Dataframe: as special kind of list
 * Tibble: as special kind of dataframe

Below I will explain the difference

## The Dataframe
The dataframe is the most widely used data structure. If you use R for doing analysis of data, this is the datstructure that you will use the most, by far!

To understand the dataframe we can best look at a named-vector.
An object in R has a value and can have attibutes. One of the most convenient attributes is the name of each element in a vector, which is collectively called the `names` attribute

```{r}
named_vector_ages <- c("Sofie" = 38, "Marc" = 45, "Marie" = 1.5)
named_vector_ages
names(named_vector_ages)
attributes(named_vector_ages)
```

When we call the `attributes` function on our named vactor, we get the names ("Walter", "Marc") of each element of the vector.
Other attributes can be the class and the dimensions and any user-defined attributes.

**Remember "Tidy data!"**   

## Subsetting vectors
To subset vectors we can use the square brackets `[]` in R.
Some examples:
```{r}
numbers <- c(1:10)

## get the 1st 
numbers[1]

## get the 1st and 3rd element
numbers[c(1,3)]

## get the second element three times
numbers[c(2,2,2)] #or
numbers[rep(2, times = 3)]

## remove the 5th element
numbers[-5]

## you can go crazy - but expect the unexpected
numbers[c(1:3, c(1,2,(c(3))))]
```

```{r, eval=FALSE}
## you can't mix positive and negative indices
numbers[c(-5,5)]
```

## The dataframe
Now that we understand the named vector we can expand this idea by creating several vectors with elements. These are then as the columns of the dataframe. One prerequisite of a `dataframe` is that all columns must have equal `length`. Individual columns of a datframe can have different data types. Remember that a vector can only contain one datatype. 
```{r}

## indivual vectors
age = c(24, 27, 19, 34)
sex = c("F","F","M", "M")
weight = c(64, 55, 80, 70)
given_names = c("Christa", "Suzan", "Matt", "John")

people_df <- data.frame(
  age,
  sex,
  weight,
  given_names
)

head(people_df)
``` 

## Or the 'tidyverse' way
```{r}
people_tbl <- tibble::tibble(
  age,
  sex,
  weight,
  given_names
)
people_tbl
```

## Viewing the contents of a dataframe
```{r, eval=FALSE}
summary(people_df)
table(people_df)

## gives the content of the data frame
head(people_df) 			
names(people_df) 
str(people_df)

## gives the content of the variable "age" from the data frame
people_df$age 
tibble::glimpse(people_tbl)

``` 

## Using index on dataframes
Using the index "[]" on a dataframe is a bit tricky. The dataframe always consists of rows and columns. Indexing a dataframe goes like:

`dataframe[row number(s), column number(s)]`

```{r, eval=FALSE}
## first element of this vector
people_df$age[1] 	

## content of 2nd variable (column) 
people_df[,2]

## content of the 1st row
people_df[1,] 	  
                  # multiple indices
people_df[2:3, c(1,3)] # remember to use c (2nd and 3rd row, 1st and 3rd column)

```

## Lists

The dataframe and the list are the most widely used datastructures when considering experimental data. Where a dataframe is constricted to the columns having all equal length, the elements of a list can be of different length and type. This is what makes the list the most flexible and sometimes the most agonizing datastructure in R.
The elemements of a list can have names, or not. Below we create a list with named elements.

## Create a List
```{r}
lst <- list(
  first_names = c(male = "Fred", 
                  female = "Mary"), 
                  no.children = 3, 
                  child_ages=c(4,7,16),
  child_names = c("Suzy", 
                  "Marvin", 
                  "Jane"), 
  address = c("Pandamonium Alley 114, Chaosville"),
  marital_status = TRUE)

lst
```

## Accessing items in a list
```{r}
# number of elements in the list
length(lst) 

lst[1] %>% class # 1st element of list as a list
lst[[1]] # 1st element of List
lst[[3]][2] # second item of third element
names(lst) # named elements in this list
lst$child_names # pull "named" elements from a list using 
lst$address

#`$` operator
```

## `str()` also gives you the structure of a list
```{r}
str(lst) # display structure of lst
``` 
The new RStudio Interface also enables interactive exploration of R-objects (demo)

## Selecting single elements in a list
To select a single element from a variable in a list
```{r}
lst$child_ages[3] 
lst[[6]][2] # returns the value of the second element for your variable
``` 

## Looping over lists
```{r}
purrr::map(lst, is.na)
```

## The matrix
A matrix is a table with only numeric values. An array consists of multiple matices. Below we create a matrix. The difference between a matrix and a dataframe is that a matrix can only hold one type of data, whereas the columns of a dataframe can be of different types of data.

Here we create a numeric matrix of 100 random normals, with 20 rows and 5 columns.
```{r}
matrix_num <- matrix(data = rnorm(n = 100),
                 nrow = 20,
                 ncol = 5) %>%
  print()
```

You can put any valid R data type in a matrix, but remember, it can only hold one type of data: A matrix with CAPITALS
```{r}
matrix_chr <- matrix(data = (c(LETTERS, LETTERS)[1:30]),
                     nrow = 5,
                     ncol = 6) %>%
  print()
                     
```

A matrix with logicals
```{r}
matrix_lgl <- matrix(data = rep(c(TRUE, FALSE, TRUE), times = 20),
                     ncol = 5) %>%
  print()
```

A matrix is basically a vector, with attributes that contain it's dimensions
```{r}
matrix_num[1:10]
attributes(matrix_num)
```

## Import data into R
```{r}
library(tidyverse)
path_to_gender_age_data <- here::here("data", "gender.txt")
gender_age <- read_delim(path_to_gender_age_data,
                         delim = "/")
```

## read_csv
CSV is a format of a data file that uses commas or semicolons as separators for the columns.

```{r}
library(readr)
skin <- read_csv(here::here("data", "skincolumns.csv")) 
skin %>% head(3)
```

## Inspecting the skin dataframe
```{r, eval=FALSE}
head(skin)	 # content of the data frame
dim(skin)
attributes(skin)
summary(skin)
## ?read_csv 	 # help on the function
```

## `skin` dataset contains an NA, some functions do not work with NAs:
```{r}
mean(skin$`Genotype A`)
mean(skin$`Genotype B`)

# to remove the NA, take care: consider leaving NAs in and use arguments like na.rm = TRUE
skin_noNA <- na.omit(skin)
mean(skin_noNA$`Genotype B`)
``` 

## Let's clean up the workspace
```{r, eval = FALSE }
rm(list=ls())
root <- find_root_file(criterion = is_rstudio_project)
## Note: never use this in code that is meant for others!!!
```

*_The above is an effective way to clear all the items in the Global Environment, but is is not very friendly to use this in code you share with others: can you think of why?_*

## Number notations and rounding

## Scientific notations 
```{r}
big_numbers <- rnorm(10, mean = 10000000, sd = 2000000)
big_numbers %>% formatC(format = "e", digits = 5)

many_digits <- c(2.55858868688584848)
round(many_digits, digits = 3)
sqrt(many_digits * 1000 /200 * 2^6)  

```

## Rounding numbers
```{r}
small_numbers <- runif(10, min = 0.001, max = 0.1) %>% print()
small_numbers %>% round(digits = 2)
```

## Generate R-code from this Rmd
```{r, eval=FALSE}
knitr::purl("hands_ons/part1_intro_to_rstudio_and_r.Rmd", output = "./hands_ons/part1_intro_to_rstudio_and_r.R")
```

## Statistics and R {.build}

 * Nowadays R is a versatile language and can be used for various applications, but it was originally designed as a language for performing statistical analyses 
 * This is why very many statistical applications are available for R
 * To start learning about statisics and R, I can highly recommend the book "Discovering Statistics Using R" by Dr. Andy Field:
 
https://uk.sagepub.com/en-gb/eur/discovering-statistics-using-r/book236067%20

## For more on learning R, start e.g. with:

http://www.statsteachr.org/ or

https://www.coursera.org/learn/statistical-inference/home/welcome or

https://www.youtube.com/watch?v=ACWuV16tdhY&index=21&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU and

https://www.youtube.com/watch?v=kvmSAXhX9Hs&index=29&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU 

and the rest of Mike Marin's lectures, that are also a really great way to start learning R.
Or try datacamp.com

## EXERCISES
```{r}
options_exercises <- knitr::opts_chunk$set(echo = FALSE,
                                           warning = FALSE,
                                           error = FALSE,
                                           message = FALSE,
                                           results = 'hide',
                                           fig.show = 'hide')
```

## 1. Load example dataset
```{r, exercise_chunks, include = FALSE}
## use other example

library(readr)
simple_IO <- read_csv(here::here(
"data",
"simple-IO.csv"
))
```
 1. What are the dimensions of the `df` dataframe?
 1. What are the column names of `df`?
 1. What are the data types of the different columns of `df?`

```{r, answer_to_1}
dim(simple_IO)
ncol(simple_IO)
nrow(simple_IO)
```

## 2. Inspecting a dataframe

 1. How do we inspect `simple_IO`?
 1. How many variables does `simple_IO` have?
 1. How many rows?
 1. call `summary()` on `simple_IO`; In which variable do we observe the highest expression ratio?

## Vector types

```{r, echo=TRUE}
a <- c("a", "b", "c", NA)
b <- c(1:4)
c <- c(6:8, NA)
d <- c(1.3, 1.6, 6.7)
e <- as.integer(d)
f <- as.integer(a)
g <- as.numeric(a)

(z <- paste(a, b, c))
(zz <- paste0(a, b, c))

```


