# Lab 3C; Tidy Data {#lab3ctidydata}

## Credits
The contents of this chapter are (partly) reproduced from [@r4ds]

## Introduction

 * In this lesson, you will learn a consistent way to structure your data in R, in a framework called __tidy data__. 
 * Getting your data into this format requires some upfront work, but that work pays off in the long term. 
 * Once you have tidy data and the tidy tools provided by packages in the tidyverse and others, you will spend much less time munging data from one representation to another, allowing you to spend more time on the analytic questions at hand.

## `{tidyr}` package

 * This lesson will give you a practical introduction to tidy data and the accompanying tools in the __tidyr__ package. 

If you'd like to learn more about the underlying theory, you might enjoy the *Tidy Data* paper published in the Journal of Statistical Software, <http://www.jstatsoft.org/v59/i10/paper>.

## Prerequisites
In this chapter we'll focus on tidyr, a package that provides a bunch of tools to help tidy up your messy datasets. tidyr is a member of the core tidyverse.

```{r packages, message = FALSE}
library(tidyverse)
```

```{r, echo=FALSE}
path_to_tidyr_sticker <- file.path(
  here::here(
  "images", 
  "tidyr_sticker.png"))
knitr::include_graphics(path_to_tidyr_sticker, dpi = 80)
```

## Tidy data rules
```{r tidy-structure, echo = FALSE}
path_to_tidydata_image <- file.path(
  here::here(
  "images",
  "images",
  "tidy-1.png"))
knitr::include_graphics(path_to_tidydata_image, dpi = 250)
```

 * Each variable goes in it's own column
 * Each observation goes in a singe row
 * Each cell contains only one value

__**You will need them in (almost) every analysis!**__

## Examples 

 * This dataset concerns tuberculosis incidence in multiple countries over 2 years
 * Each dataset shows the same values of four variables *country*, *year*, *population*, and *cases*
 * Ask yourself: "How are the following tables different?"
 * Check if, and how the table complies to the tidy data rules

## Example 1
```{r}
head(table1, 6)
```

## Example 2
```{r}
head(table2, 6)
```

## Example 3
```{r}
head(table3, 6)
```

## Data can be spread across multiple files/tables
```{r}
head(table4a, 3)  # cases
head(table4b, 3)  # population
```

## One winner!
In the previous example, only `table1` is tidy. It's the only representation where each column is a variable. 

```{r} 
head(table1)
```

## Why tidy data?
Why ensure that your data is tidy? There are two main advantages:

 1. There's a general advantage to picking one consistent way of storing data. If you have a consistent data structure, it's easier to learn the tools that work with it because they have an underlying uniformity.
    
 2. There's a specific advantage to placing variables in columns because they are vectors. Most built-in R functions work with vectors of values.

## Packages from the tidyverse are designed for tidy data 

Example plot from `table1`
```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
```

## Compute cases per year
```{r}
table1 %>% 
  count(year, wt = cases)
```

## **EXERCISE 1** Visualize changes over time {-}

A) Create a plot from `table1` tuberculosis data, showing the number of cases from each country.

B) Now plot the incidence (number of cases per 10,000 people) for each country.

C) What can you conclude from the plots in exercise 1A) and 1B)

```{r, include=FALSE}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

## Gathering

When some of the column names are not names of variables, but _values_ of a variable. Take `table4a`: the column names `1999` and `2000` represent values of the `year` variable, and each row represents two observations, not one.

```{r}
table4a
```

## Gathering a set of value columns

To tidy a dataset like this, we need to __gather__ those columns into a new pair of variables. To describe that operation we need three parameters:

* The set of columns that represent values, not variables. In this example, 
  those are the columns `1999` and `2000`.
* The name of the variable whose values form the column names. I call that
  the `key`, and here it is `year`.
* The name of the variable whose values are spread over the cells. I call 
  that `value`, and here it's the number of `cases`.
  
## "gather()"  
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

## The actions "gather()" performs
```{r tidy-gather, echo = FALSE, out.width = "100%", fig.cap = "Gathering `table4` into a tidy form."}
knitr::include_graphics(file.path(
  here::here(
  "images", 
  "images", 
  "tidy-9.png")
  ))
```

## Tidy table 4b
```{r}
table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
```

## **EXERCISE 2** Gathering columns {-}
The data file "./data/messy_excel.xlsx" is not very tidy. Open the file in Excel. Discuss with you neighbour how you would tidy the data in this file, without changing anything in the Excel file. Write down your plan. try to develop a stepwise protocol that includes code-snippets and/or pseudo code.

**TIPS**
 
 1. You can define a range to be parsed in the readxl package
 1. You will need gather some columns in the data
 1. You might want to parse the file 2 times with different ranges to get all the data into R
 1. First try to parse the file as it is to see what R makes of the file and try to design (in Excel if you want) the end result that you would like to achieve.

_This is a difficult exercise and it will take you some time to figure out how to achieve this, take the time to look for additional online information_

```{r, include=FALSE}
## see demo "readxl"
```

## Joining tables
To combine the tidied versions of `table4a` and `table4b` into a single tibble, we need to use `dplyr::left_join()`
```{r, message=FALSE}
tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
joined_4a_4b <- left_join(tidy4a, tidy4b)
joined_4a_4b
```

## Spreading
Spreading is the opposite of gathering. You use it when an observation is scattered across multiple rows. For example, take `table2`: an observation is a country in a year, but each observation is spread across two rows.

```{r}
table2
```

To tidy this up, we first analyse the representation in similar way to `gather()`. This time, however, we only need two parameters:

* The column that contains variable names, the `key` column. Here, it's 
  `type`.
* The column that contains values forms multiple variables, the `value`
  column. Here it's `count`.

## "spread()"
```{r}
spread(table2, key = type, value = count)
```

## The action "spread()" performs
```{r tidy-spread, echo = FALSE, out.width = "100%", fig.cap = "Spreading `table2` makes it tidy"}
knitr::include_graphics(file.path(
  here::here(
  "images", 
  "images", 
  "tidy-8.png")
  ))
```

## "gather()" and "spread()" are complementary
As you might have guessed from the common `key` and `value` arguments, `spread()` and `gather()` are complements. `gather()` makes wide tables narrower and longer; `spread()` makes long tables shorter and wider.

## Separating and uniting

`table3` has a different problem: we have one column (`rate`) that contains two variables (`cases` and `population`). To fix this problem, we'll need the `separate()` function. You'll also learn about the complement of `separate()`: `unite()`, which you use if a single variable is spread across multiple columns.

## "separate()"

`separate()` pulls apart one column into multiple columns, by splitting wherever a separator character appears. Take `table3`:

```{r}
table3
```

## Separating rate into cases and population
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

## The action "separate()" performs
```{r tidy-separate, echo = FALSE, out.width = "75%", fig.cap = "Separating `table3` makes it tidy"}
knitr::include_graphics(file.path(
  here::here(
  "images", "images", "tidy-17.png")))
```

## Unite
`unite()` is the inverse of `separate()`: it combines multiple columns into a single column. 

Look at `?tidyrt::unite()` to see the help and examples

## Missing values

Changing the representation of a dataset brings up an important subtlety of missing values. Surprisingly, a value can be missing in one of two possible ways:

* __Explicitly__, i.e. flagged with `NA`.
* __Implicitly__, i.e. simply not present in the data.

Let's illustrate this idea with a very simple data set:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

## The stocks dataframe
```{r}
stocks
```

## There are two missing values in this dataset:

 * The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains `NA`.
 * The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

## Making implicit missing values explicit
The way that a dataset is represented can make implicit values explicit. For example, we can make the implicit missing value explicit by putting years in the columns:

```{r}
stocks %>% 
  spread(year, return)
```

## "na.rm = TRUE"
Because these explicit missing values may not be important in other representations of the data, you can set `na.rm = TRUE` in `gather()` to turn explicit missing values implicit:

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`, na.rm = TRUE)
```

## Explicit missing values "complete()"
Another important tool for making missing values explicit in tidy data is `complete()`:

```{r}
stocks %>% 
  complete(year, qtr)
```


## The "complete()" function
`complete()` takes a set of columns, and finds all unique combinations. It then ensures the original dataset contains all those values, filling in explicit `NA`s where necessary.

## Forward carrying
Sometimes when a data source has primarily been used for data entry, missing values indicate that the previous value should be carried forward:

## Carry-forward example
```{r}
(treatment <- tribble(
  ~ person,            ~ treatment, ~response,
  "Hans Anders",       1,           7,
  NA,                  2,           10,
  NA,                  3,           9,
  "Alber Heijn",       1,           4,
  NA,                  2,           8,
  NA,                  3,           12
))
```

## Filling in the 'missing values'
You can fill in these missing values with `fill()`. It takes a set of columns where you want missing values to be replaced by the most recent non-missing value (sometimes called last observation carried forward).

```{r}
treatment %>% 
  fill(person)
```


## **EXERCISE 3** Inspecting tidyness {-}

3A) Describe how the variables and observations are organised in each of the sample tables: 
`table1`, `table2`, `table3`, `table4a`, `table4b`.

3B) Compute the `rate` for `table2`, and `table4a` + `table4b`. 

**STEPS**

You will need to perform four operations:

 1.  Extract the number of TB cases per country per year.
 2.  Extract the matching population per country per year.
 3.  Divide cases by population, and multiply by 10000.
 4.  Store back in the appropriate place.
 5.  Sort the data according year (ascending)
 
**TIPS**
 
 - You will need {tidyr} and {dplyr}
 - You will also need `mutate()` to calculate the rate
 - Use the pipe `%>%` 
 - Rember `gather()` collects observations in columns into variable columns (from wide to stacked) and `spread()` unstacks variables into variable columns (from stacked to wide)
 - Table 4a contains the cases, table 4b contains the populations
 - For joining the tables 4a and 4b you can use the `dplyr::left_join()` function. See `help("dplyr")` or `?dplyr` for more information.

table2
```{r, include=FALSE}

library(tidyr)
library(dplyr)
table2 <- table2
table2_tidy <- table2 %>%
  spread(key = "type", value = "count") %>%
  mutate(rate = (cases/population)*1000) %>%
  arrange(year)
table2_tidy
```

table 4a + 4b
```{r, include=FALSE}
## ANSWER 4a + 4b
table4a <- tidyr::table4a
table4b <- tidyr::table4b
table4a
table4b

table4a_tidy <- table4a %>%
  gather(`1999`:`2000`, key = "year", value = "cases")

table4b_tidy <- table4b %>%
  gather(`1999`:`2000`, key = "year", value = "population")

table4_union <- left_join(table4a_tidy, 
                          table4b_tidy, 
                          by = c("country", "year")) %>%
  mutate(rate = (cases/population)*1000) %>%
  arrange(year)

table4_union
```

Which representation is easiest to work with? Which is hardest? Why?

## EXERCISE 2: Plotting tidy data {-}

2A)  Recreate the plot showing change in cases over time using `table2`
instead of `table1`. What do you need to do first?

```{r, include=FALSE}
## answer 2a
library(ggplot2)

cases_over_time <- table2 %>%
  spread(key = "type", value = "count") %>%
  mutate(rate = (cases/population)*1000) %>%
  arrange(year) %>%
  ggplot(aes(x = as.factor(year), y = cases)) +
  geom_point(aes(colour = country)) +
  geom_text(aes(label = country), hjust = -0.2, vjust = -0.7) +
  geom_line(aes(group = country, colour = country)) +
  theme(legend.position="none") +
  expand_limits(y = c(0, 250000)) +
  xlab("Year") +
  ylab("Number of Cases")
                 
cases_over_time

```

2B) Why does the code below fail, can you fix it?
```{r, error = TRUE, echo=TRUE}
    table4a %>% 
      gather(1999, 2000, key = "year", value = "cases")
```

## EXERCISE 3: Wide to long; Spreading and Gathering {-}

3A)  Tidy the simple tibble below. Do you need to spread or gather it?
```{r, echo=TRUE, include=FALSE}

    preg <- tribble(
      ~pregnant, ~male_count, ~female_count,
      "yes",      NA,        18,
      "no",       20,        12,
      "no",       34,        78,
      "yes",      NA,        23
)

```

```{r, include=FALSE}
## answer 3a
preg_tidy <- preg %>%
   gather(male_count:female_count, 
          key = "count", 
          value = number)
preg_tidy
```

3B) Calculate the percentage of females that are pregnant, of all females in the dataset. 
```{r, include=FALSE}
## answer 3b
names(preg_tidy)

preg_females <- preg_tidy %>%
  dplyr::filter(pregnant == "yes" & count == "female_count") %>%
  summarise(females = sum(number)) 
preg_females

females <- preg_tidy %>%
  dplyr::filter(count == "female_count") %>%
  summarise(females = sum(number)) 
females

((preg_females/females)*100)

```

3C) How many males are there in the dataset? How many of them are pregnant ;-)
```{r}
## answer 3c
names(preg_tidy)

males <- preg_tidy %>% na.omit() %>%
  dplyr::filter(count == "male_count") %>%
  summarise(males = sum(number)) 
males

preg_males <- preg_tidy %>% na.omit() %>%
  dplyr::filter(pregnant == "yes" & count == "male_count") %>%
  summarise(females = sum(number)) 
preg_males

## of course there are nog pregnant males

```

## EXERCISE 4: Addiotional operations {-}

4A) What do the `extra` and `fill` arguments do in `separate()`? Experiment with the various options for the following two toy datasets.
    
```{r, eval = FALSE}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
separate(x, c("one", "two", "three"))
```

4B) Compare and contrast the functions `separate()` and `extract()`.  Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

## EXERCISE 5: A Case {-}

5A) Tidy the following dataset: "./data/messydata_5a.xlsx"
The data represents Mariages during the four seasons for a number of years in regions in Germany. 

Note: Germany has 16 states. When you load the data you will notice that the coding of the factor variable "State" is not consistent. You will need to correct tihs coding in order for the data to be tidy.
Use the function from {forcats} `forcats::fct_recode()` to achieve this. This file "./data/germany_states.txt" contains information on the states. How can you load this file into R?


**TIPS**
 
 - Use the code below to read all the sheets of the excel file, look at the help function first.
 - Which parameter of `read_excel()` controls the sheets to read? 
 - To read multiple sheets, a for loop is used
 - The Excel file contains 7 sheets 
 - Check spelling and values: are they consistent. 
 - Read the chapter "
 - Think about `gather()` or `spread()`, which do you need?
 - To convert a list of dataframes to one dataframe look at:
 the help for the function `dplyr::bind_rows()`. You will have to use this to convert the list of sheets into one dataframe.
- The first two sheets of the Excel file contain mistakes and errors

### Read data {-}
```{r, echo=TRUE}
library(readxl)

# path to data file
path_to_messy_mariages <- file.path(
  here::here(
    "data", 
    "messydata_5a.xlsx"))

# load messy data in empty list with loop and read_excel
messy_mariages <- list()

for(i in 1:7){
  messy_mariages[[i]] <- read_excel(path = path_to_messy_mariages, 
                                    sheet = i)
  messy_mariages[[i]]$timestamp <- i
}


```


```{r, include=FALSE}
# loading required packages
library(tidyverse)
library(readxl)
library(stringr)
library(forcats)

# path to data file
path_to_messy_mariages <- file.path(
  here::here(
    "data", 
    "messydata_5a.xlsx"))

# load messy data in empty list with loop and read_excel
messy_mariages <- list()

for(i in 1:7){
  messy_mariages[[i]] <- read_excel(path_to_messy_mariages, 
                                    sheet = i)
  messy_mariages[[i]]$timestamp <- i
}

messy_mariages
messy_sheet_1 <- messy_mariages[[1]]
messy_sheet_1 <- messy_sheet_1[c(2:17), c(4:9)]
## change names of first sheet to match others
names(messy_sheet_1) <- c(names(messy_mariages[[3]]))                     
## overwrite old sheet with new, with correct names
messy_mariages[[1]] <- messy_sheet_1 


# bind list to data frame
messy_mariages_df <- do.call(rbind.data.frame, messy_mariages)

# load the state information from "./data/germany_states.txt"
# Does the file contain headers?
germany_states <- readr::read_tsv(
  file = 
    here::here(
    "data",
    "germany_states.txt"),
  col_names = FALSE)
# restructuring data
## gathering variables into columns
restructured_mariages <- messy_mariages_df %>% 
  tidyr::gather(spring:winter, 
                key = "season", 
                value = "number_mariages")
  
restructured_mariages$timestamp <- as.factor(
  restructured_mariages$timestamp)

levels(restructured_mariages$timestamp)

tidy_mariages <- restructured_mariages %>%
# change timestamps to actual Years
## use forcats::recode
mutate(year = forcats::fct_recode(timestamp,
                                  "1990" = "1",
                                  "2003" = "2",
                                  "2010" = "3",
                                  "2011" = "4",
                                  "2012" = "5",
                                  "2013" = "6",
                                  "2014" = "7"))
  
# look at the restructed data
tidy_mariages

# set 'names' tolower
names(tidy_mariages) <- tolower(names(tidy_mariages))
names(tidy_mariages)

# have a closer look at the Statenames (convert to factor)
tidy_mariages$state <- as_factor(tidy_mariages$state)
levels(tidy_mariages$state)

# relabel factor levels
tidy_mariages <- tidy_mariages %>%
  mutate(state = fct_recode(state,
                      "Baden-W端rttemberg" = "BaW端",
                      "Hessen" = "Hesssen",
                      "Hessen" = "Hesse"))
                      
levels(tidy_mariages$state)


# save tidy data as csv
path <- file.path(
  here::here(
  "data", "tidy_mariages.csv"))

write_csv(tidy_mariages, path = path)

```

5B) Start analyzing your tidy data

 - How many marriages are there per year in each German "Bundesland"?
 - Visualize your findings in a plot
 
```{r, include=FALSE}
marriageperyear <- tidy_mariages %>% 
  group_by(year) %>% # groups data by year
  dplyr::summarise(total = sum(as.numeric(number_mariages))) # adds up the values per year

marriageperyear %>%
  ggplot(aes(x = year,
             y = total)) +
  geom_point() +
  geom_line(aes(group = 1))

```

5C) Visual analysis
Compare the marriages of 1990, 2003 and 2014 and colour them by season

**TIPS:**
 - Filter data by the years 1990, 2003 and 2014
 - Group it by year and season and add up the values

```{r, include=FALSE, eval=FALSE}
marriageperyearnseason <- tidy_mariages %>%
  dplyr::filter(year %in% c("1990", "2003", "2014")) %>%
  group_by(year, season) %>% 
  mutate(number_mariages = as.numeric(number_mariages)) %>%
  dplyr::summarise(value = sum(number_mariages)) 

# ggplot the filtered data
ggplot(data = marriageperyearnseason, 
       aes(x = year, y = value)) +  
geom_bar(stat = "identity") +
theme_minimal() +
xlab("Year") + 
scale_y_continuous(name="Marriages", 
                   # specify aesthetics of y-axis labels
                   labels=function(x) format(x, big.mark = "'", scientific = FALSE)) + 
guides(fill=guide_legend(title="season", reverse = T)) +
ggtitle("Marriages per Year and season") 

# Compare the number of marriages in 2014 per state and colour them by season
# filter data by the year 2014 and group it by season and state, add up the values
marriageperstate14 <- tidy_mariages %>%
  dplyr::filter(timestamp %in% c(2014)) %>%
  group_by(number_mariages, state) %>% 
  dplyr::summarise(total = sum(as.numeric(number_mariages))) 

# stacked barplot with ggplot (incl. dashed meanline)
meanlabel <- (sum(marriageperstate14$number_mariages)/16)-2000 # y-coordinate of label for meanline

ggplot(data = marriageperstate14, aes(x = State, y = value, 
                                      fill = factor(variable), order = variable)) +  
  geom_bar(stat = "identity") +
  theme_minimal() +
  xlab("State") + 
  scale_y_continuous(breaks = seq(0,85000, 10000), name="Marriages", 
                     labels=function(x) format(abs(seq(0,85000,10000)), 
                                               big.mark = "'", scientific = FALSE)) +
  scale_x_discrete(labels=c("BaW端","Bay","Ber","Bra","Bre","Ham","Hes","Meck","Nie","Nor",
                            "Rhe","Saa","Sac","SaAn","Sch","Th端")) + # change x-axis labels
  theme(axis.text.x = element_text(angle = - 50, vjust = 0.9, hjust = 0.1)) +
  guides(fill=guide_legend(title="season", reverse = T)) +
  ggtitle("Marriages per state and season \n 2014") +
  geom_hline(aes(yintercept = (sum(value)/16)), color = "black", linetype = "dashed", 
             size = 0.5) + # add meanline
  annotate("text", x = 10, y = meanlabel, label="Mean", color="black", 
size = 4, hjust = 1) # add text to mean line


```

## Case Study {-}

To finish off the chapter, let's pull together everything you've learned to tackle a realistic data tidying problem. The `tidyr::who` dataset contains tuberculosis (TB) cases broken down by year, country, age, gender, and diagnosis method. The data comes from the *2014 World Health Organization Global Tuberculosis Report*, available at <http://www.who.int/tb/country/data/download/en/>.

There's a wealth of epidemiological information in this dataset, but it's challenging to work with the data in the form that it's provided:

```{r}
who
```

This is a very typical real-life example dataset. It contains redundant columns, odd variable codes, and many missing values. In short, `who` is messy, and we'll need multiple steps to tidy it. Like dplyr, tidyr is designed so that each function does one thing well. That means in real-life situations you'll usually need to string together multiple verbs into a pipeline. 

The best place to start is almost always to gather together the columns that are not variables. Let's have a look at what we've got: 

* It looks like `country`, `iso2`, and `iso3` are three variables that 
  redundantly specify the country.
  
* `year` is clearly also a variable.

* We don't know what all the other columns are yet, but given the structure 
  in the variable names (e.g. `new_sp_m014`, `new_ep_m014`, `new_ep_f014`) 
  these are likely to be values, not variables.

So we need to gather together all the columns from `new_sp_m014` to `newrel_f65`. We don't know what those values represent yet, so we'll give them the generic name `"key"`. We know the cells represent the count of cases, so we'll use the variable `cases`. There are a lot of missing values in the current representation, so for now we'll use `na.rm` just so we can focus on the values that are present.

```{r}
who1 <- who %>% 
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)
who1
```

We can get some hint of the structure of the values in the new `key` column by counting them:

```{r}
who1 %>% 
  count(key)
```

You might be able to parse this out by yourself with a little thought and some experimentation, but luckily we have the data dictionary handy. It tells us:

1.  The first three letters of each column denote whether the column 
    contains new or old cases of TB. In this dataset, each column contains 
    new cases.

1.  The next two letters describe the type of TB:
    
    *   `rel` stands for cases of relapse
    *   `ep` stands for cases of extrapulmonary TB
    *   `sn` stands for cases of pulmonary TB that could not be diagnosed by 
        a pulmonary smear (smear negative)
    *   `sp` stands for cases of pulmonary TB that could be diagnosed be 
        a pulmonary smear (smear positive)

3.  The sixth letter gives the sex of TB patients. The dataset groups 
    cases by males (`m`) and females (`f`).

4.  The remaining numbers gives the age group. The dataset groups cases into 
    seven age groups:
    
    * `014` = 0 -- 14 years old
    * `1524` = 15 -- 24 years old
    * `2534` = 25 -- 34 years old
    * `3544` = 35 -- 44 years old
    * `4554` = 45 -- 54 years old
    * `5564` = 55 -- 64 years old
    * `65` = 65 or older

We need to make a minor fix to the format of the column names: unfortunately the names are slightly inconsistent because instead of `new_rel` we have `newrel` (it's hard to spot this here but if you don't fix it we'll get errors in subsequent steps). You'll learn about `str_replace()` in [strings], but the basic idea is pretty simple: replace the characters "newrel" with "new_rel". This makes all variable names consistent.

```{r}
who2 <- who1 %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
who2
```

We can separate the values in each code with two passes of `separate()`. The first pass will split the codes at each underscore.

```{r}
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

Then we might as well drop the `new` column because it's constant in this dataset. While we're dropping columns, let's also drop `iso2` and `iso3` since they're redundant.

```{r}
who3 %>% 
  count(new)
who4 <- who3 %>% 
  select(-new, -iso2, -iso3)
```

Next we'll separate `sexage` into `sex` and `age` by splitting after the first character:

```{r}
who5 <- who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)
who5
```

The `who` dataset is now tidy!

I've shown you the code a piece at a time, assigning each interim result to a new variable. This typically isn't how you'd work interactively. Instead, you'd gradually build up a complex pipe:

```{r, results = "hide"}
who %>%
  gather(code, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(code = stringr::str_replace(code, "newrel", "new_rel")) %>%
  separate(code, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

### Exercises

1.  In this case study we set `na.rm = TRUE` just to make it easier to
    check that we had the correct values. Is this reasonable? Think about
    how missing values are represented in this dataset. Are there implicit
    missing values? What's the difference between an `NA` and zero? 

1.  What happens if you neglect the `mutate()` step?
    (`mutate(key = stringr::str_replace(key, "newrel", "new_rel"))`)

1.  I claimed that `iso2` and `iso3` were redundant with `country`. 
    Confirm this claim.

1.  For each country, year, and sex compute the total number of cases of 
    TB. Make an informative visualisation of the data.



