# Introduction {#intro}

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 6, fig.height = 4)
```

## Exploratory Data Analysis

(ref:eda) The process of Exploratory Data Analysis (EDA) is not a formal strict process. It involves interative cycles of `loading`, `cleaning`, `wrangling`, `visualizing`, `communicating` data and patterns in the data. The process of EDA is directed towards gaining insight in the data in basically any way you can. It can involve a number of statical approaches that are sometimes collectively called IDA (Initial Data Analysis), which is the the description and check if undelying assumptions for any formal statitical modelling are met. Usually the formal statistical inference is out of scope of the EDA process, although we will see examples in which statical modelling can help us perform EDA better. Because it is a not-so-strict iterative process there is no formal manual on which steps to perform in the EDA process. You could consider doing EDA more as being in a certain state-of-mind. To help you overcome this rather abstract way of looking at this process, several authors have created a check list to aid performing EDA in a more structured way. Here I will go over such a check list, but bare in minf that is only an aid and no formal manual: __When doing EDA you should always keep an open mind to deviate from the checklist, skip a check-box or add one yourself.__

(ref:eda)


## EDA checklist 

"If a checklist is good enough for pilots to use before and after every flight, it’s good enough for data scientists to use with every dataset." A quote from Daniel Bourke on [Towards Data Science](https://towardsdatascience.com/a-gentle-introduction-to-exploratory-data-analysis-f11d843b8184)

When perfoming EDA consider:

 1. What question(s) are you trying to solve (or prove wrong)?
 1. Which information do you need and can you come up with a plan to answer the question(s)
 1. What kind of data do you have and how do you treat different types?
 1. What’s missing from the data and how do you deal with it?
 1. Where are the outliers and why should you care about them?
 1. How can you add, change or remove features to get more out of your data?
 1. Do you need additional data from other sources to relate to the dataset under scrutiny?
 1. Are underlying statistical assumptions met / how is data distribution looking?
 1. What (exploratory) models apply or fit well to the data?
 1. What is the underlying (experimental) design?
 1. Is there multi-colinearity?

I will go over each item in more detail in Chapter \@ref(lab5eda). If you want to get on with a first example to have some practice go directly to Chapter \@ref(edacase). If you have nog programming experience with R, I recommend going over Chapter \@ref(lab1aintrorstudio) and \@ref(lab1bintror) first, before you start with the case example in Chapter \@ref(edacase).

## The `{tidyverse}` packages

The `{tidyverse}` [@tidyverse] is a suite of packages that together form a nice collection of tools to perform different tasks and analyses in R. A complete documentation website to the `{tidyverse}` is maintained [here](https://www.tidyverse.org/)

Many of these packages have multiple contributors but one name cannot be left unmentioned in a paragraph about the tidyverse and that is the name of Dr. Hadley Wickham, Chief Scientist at RStudio, and an Adjunct Professor of Statistics at the University of Auckland, Stanford University, and Rice University. Hadley is one of the driving forces behind innovation in the R-community when in comes to writing more robust, more reproducible and more `human-readable` computer code. 

The `{tidyverse}` is meant as a collection of helpful tools for doing data science with R, but it is not meant as a stand-alone suite to be used in isolation. Connecting tidyverse to other packages is encouraged and also feasible and most of the times even necessary to get the job done.

To review some of the tidyverse principles and programming examples in this course the participant is encouraged to read the very comprehensive and complete book written by Garret Grolemund and Hadley wickham called "R for Data Science". This book is freely available as a bookdown website at https://r4ds.had.co.nz/

To download the materials and if you want to recompile it from source: https://github.com/hadley/r4ds. The book is also available in print version, would you prefer that.

loading the tidyverse can be achieved by running this code:
```{r}
## if not installed already: install.packages("tidyverse")
library(tidyverse)
```

## Inference and Modelling
This course will not be a statistics course!! So you can relax a bit more now. BUT: we will need som statistical principles and insights to succesfully exploit the complete EDA process. EDA is all about providing insignt into data, you most of the time know very little about on beforehand. Statistics can be very helpfull to find and assess patterns in a more formal way. During this course we will be avoiding in-depth explanations about statitical models and we will use a little formulas as possible. Most of the time, we focus on exploring assumptions that are a prequisite for perfoming statistics. 

Sometimes, we will also looks at more complex examples like for example regressions, regression tree methods like Random Forest or Boosting and Analysis of Variance methods. This is mainly meant as demonstrations how to build upon EDA with modelling and to illustrate how model performace and inferential results can be obtained and read.

## R: A Language and Environment for Statistical Computing 
The following section was taken from the Wikipage on R at: https://en.wikipedia.org/wiki/R_(programming_language). 

If you are using R and want to refer it in your reference list you can get the full citation like:
```{r}
citation()
```
Users of LaTeX can also use the BibTex entry to create a reference list.

### History of R

R is an implementation of the S programming language combined with lexical scoping semantics, inspired by Scheme. S was created by John Chambers in 1976, while at Bell Labs. There are some important differences, but much of the code written for S runs unaltered.

R was created by Ross Ihaka and Robert Gentleman[17] at the University of Auckland, New Zealand, and is currently developed by the R Development Core  Team (of which Chambers is a member).[18] R is named partly after the first names of the first two R authors and partly as a play on the name of S. The project was conceived in 1992, with an initial version released in 1995 and a stable beta version in 2000 [text from wiki](https://en.wikipedia.org/wiki/R_(programming_language).

### Statistical features

R and its libraries implement a wide variety of statistical and graphical   techniques, including linear and nonlinear modeling, classical statistical  tests, time-series analysis, classification, clustering, and others. R is  easily extensible through functions and extensions, and the R community is  noted for its active contributions in terms of packages. Many of R's standard functions are written in R itself, which makes it easy for users to follow the algorithmic choices made. For computationally intensive tasks, C, C++, and Fortran code can be linked and called at run time. Advanced users can write C, C++,[23] Java,[24] .NET[25] or Python code to manipulate R objects directly.[26] R is highly extensible through the use of user-submitted packages for specific functions or specific areas of study. Due to its S heritage, R has stronger object-oriented programming facilities than most statistical computing languages. Extending R is also eased by its lexical scoping rules.[27]

Another strength of R is static graphics, which can produce publication-quality graphs, including mathematical symbols. Dynamic and interactive graphics are available through additional packages.[28]

R has Rmd, its own LaTeX-like documentation format, which is used to supply comprehensive documentation, both online in a number of formats and in hard copy.[29] 

For more details on R itself review the Wiki page or look at: https://cran.r-project.org/

## Getting help in R
As mentioned above R is known for its elaborate documentation. One of the biggest reasons to prefer R above other data science languages like e.g. Python or Perl is that documentation in R is obligatory for each package (hosted on CRAN or BIOCONDUCTOR). This makes finding useful functions and packages and using them in the right context more easy. Let's look at an example:

Consider that you would like to read a csv file into R. Here we will use the standard base-R function `read.csv()` to illustration how the help function works.

To get information on a function type `help("function-or-package-name")` or
`?function-or-package-name` in the Console

```{r, eval=FALSE}
help()
```


## Resources for learning
There are many resources for learning R and doing Data Science with R. One was already mentioned and that is the R for Data Science Book [@r4ds].

A non-exhaustive list is provided in \@ref(resources) and is meant to head you in the right direction. Finding the proper learning reasource that will solve you specific 'Analytics' issue is up to you and part of the fun. Google is defenitively a good mentor to guide you through this process! 

## **EXERCISE** {-}
Using the function `installed.packages()` and running the code below. Try to figure out how many packages are installed on the RSudio server instance that you have logged onto.

Copy-paste the code below into your R Console and run the code

```{r, results='hide'}
library(tidyverse)
pkgs <- installed.packages() %>%
  as_tibble() 
pkgs
```

## --- END EXERCISE ---

## A plot on the packages
Each package has a so-called DESCRIPTION file that contains information on the title, the author(s), the aim of the package and which dependencies (other R-packages) it needs to function. The imports field is an important files that displays the essential dependencies without which the package is not able to function. Let's try to find the package that has the most dependencies listed in the imports field. 
```{r}
names(pkgs)

pkgs[1:3, "Imports"]


pkgs %>%
  select(Package, Imports) %>%
  mutate(Imports_chr = as.character(Imports)) %>%
  mutate(
    n_depends = 
      str_split(Imports_chr, pattern = ",")) %>%
 # mutate(
#    n_depends = map(n_depends, flatten)) %>%
  mutate(
    n_depends = map_int(n_depends, length)) %>%
  arrange(desc(n_depends)) %>%
  dplyr::filter(n_depends >10) %>%
  ggplot(aes(x = reorder(Package, n_depends), y = n_depends)) +
  geom_point() +
  coord_flip()

  
```

## Course Contents {#coursecontents}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Mining Course
This course is about "Exploratory Data Analysis and Initial Data Analysis"

[Wikipedia definition](https://en.wikipedia.org/wiki/Exploratory_data_analysis)

"In statistics, exploratory data analysis (EDA) is an approach to analyzing data sets to summarize their main characteristics, often with visual methods. A statistical model can be used or not, but primarily EDA is for seeing what the data can tell us beyond the formal modeling or hypothesis testing task. Exploratory data analysis was promoted by John Tukey to encourage statisticians to explore the data, and possibly formulate hypotheses that could lead to new data collection and experiments. EDA is different from initial data analysis (IDA),[1] which focuses more narrowly on checking assumptions required for model fitting and hypothesis testing, and handling missing values and making transformations of variables as needed. EDA encompasses IDA"

## Course aims
 
  - EDA is not a formal procedure, getting the right mind set
  - Learn tools in R to GET, CLEAN, EXPLORE and MODEL data
  - Acquire R skills for the complete EDA cycle; object oriented and functional programming  
  - Visualize data
  - Explore assumptions (IDA)
  - Using R for Reproducible Research 
  
## Course contents
Each course masterclass or so-called 'lab' will carry forward a specific theme related to Exploratory Data Analysis. In this course you will exclusively work with the Programming Environment `R` [@R-base] in the Integrated Development Environment `RStudio` [@R-rstudio]. There will be no need to install any software on your laptop, because we will be using a cloud computing solution. This offers flexibility and speed. I will show you where to get the required software. In case that you would be wanting to install your own environment locally.   
  
In this course you will get in-depth knowledge on how to use R in conjunction with RStudio to `IMPORT`, `INVESTIGATE`, `CLEAN`, `VISUALIZE`, `EXPLORE`, `MODEL` and `COMMUNICATE` data and conclusions about data. 
  
To this end, I divided an number of logically connected topics together in seven `labs`. Each lab has several interactive cycles of theory and exercises. During each lab, I will explain a small amount of topics after which the course participants will have the opportunity to practice with exercises.   
  
## BYOD; Bring Your Own Data  
  
During this course you will have the opportunity to bring your own data as case example. Please think about which data you will be able to (freely) share with me and your fellow paticipants. 

## Lab Contents
Below, I will shortly summarize what we will be covering in each lab (seven in total for the complete course).

### Lab 1a: Introduction to RStudio 
Chapter \@ref(lab1aintrorstudio)

 1. Getting the course materials from github.com
 1. Creating objects in R
 1. RStudio introduction

### Lab 1b: Introduction to R  
Chapter \@ref(lab1bintror)

 1. Functional and object oriented programming
 1. Functions
 1. Object Class; vectors, dataframes, lists
 1. Vector types
 1. Getting help
 1. Plots
 1. Data examples

### Lab 1c: Dataframes, lists and matrices  
Chapter \@ref(lab1crecursivevectors)

 1. Dataframes
 1. Lists
 1. Matrices

### Lab 2: Visualize & Explore Data
Chapter \@ref(lab2viz)

 1. Visualizations demo
 1. Build in datasets 
 1. Using the grammar of graphics with `{ggplot2}`
 1. Plot types
 1. `geom_...` and `aes()`
 1. Adding dimensions to a plot 
 1. Solving overplotting
 1. Plot annotation and labels
 1. Saving plots
 1. Examples of `{shiny}` apps

### Lab 3a: Data Wrangling  
Chapter \@ref(lab3awrangling)

 1. Using `{dplyr}` for data wrangling
 1. The `{dplyr}` verbs
 1. Data wrangling in combination with `{ggplot2}`

### Lab 3b: Functional Programming 
Chapter \@ref(lab3bfunctions)

 1. What is a function
 1. Writing function
 1. Function documentation
 1. Loops and `map` family of functions
 1. Scripts and the `source` function
 1. Why write function
 1. Function elements (name/arguments, body, return)
 1. Function documentation, Roxygen comments
 1. Writing error / warning /  messages
 
### Lab 3c: Tidy data
Chapter \@ref(lab3ctidydata)
 
### Lab 4a: Importing data 
Chapter \@ref(lab4aimportingdata)

 1. Paths
 1. Importing data functions
 1. Excel and other 'foreign' formats
 1. Reading from a zipfile/url

### ### Lab 4b: Getting Open Data  
Chapter \@ref(lab4bopendata)

 1. Open data sources
 1. Finding data
 1. Importing data into R (.csv, .tsv, .txt, .json, .xml, .xls(x)
 1. Using APIs (Twitter, Kaggle, Google, CBS)
 1. Accessing databases

### Lab 5: Exploratory Data Analysis _or_ The PPDAC Cycle
Chapter \@ref(lab5eda)

 1. The process of EDA
 1. Problem-Plan-Data-Analysis-Conclusion (PPDAC) Cycle
 1. Missingness
 1. Distibutions
 1. Finding patterns 
 1. Graph types
 1. Multidimensional data
 1. Principal Components
 
### Lab 6a: Exploring Assumptions & Models  
Chapter \@ref(lab6amodels)

 1. Why Assumptions?
 1. Distributions (Gaussian)
 1. Regression
 1. Model output ( `{broom}`)
 
### Lab 6b/c: Statistics with R (resampling & simple regression) 
Chapter \@ref(lab6bmanymodels)
 
### Lab 7: Machine learning with R
 
## Exercises
Each masterclass/lab contains a number of exercises. Most lessons have exercises in between the demos and explanations. Usually at the end of each lessons, more exercises are included.

In order for you to get the most out of this course it is important to go over each exercise and try to solve each challenge strictly from within RStudio. 
The easiest way to make an exercise is to copy the text of the exrcise from the Bookdown website into an RMarkdown file and place R-code chunks in between containing the answers or to use the Rmd source documents from Github directly. Store each Rmd file for each lesson seperately in a convenient folder called "exercises" in your course project folder.

During the classes we will also look at the solutions to the exercises together.

The R code containg the demos and the answers will be made available for each lab in the github repository: https://github.com/uashogeschoolutrecht/exploratory_data_masterclasses

## Getting Started {#gettingstarted}

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4,
                      warning = FALSE,
                      error = FALSE, message = FALSE)
```

## Packages
```{r}
library(tidyverse)
```


## Preface

This is an R-Markdown document. It is an example of literate programming and it illustrates how we can combine code, text and output. It addresses the issue "Write code for humans, not for computers" very nicely.

Code can be recognized by a code chunck:

_start code chunk_

` ```{r} `

`library(tidyverse)`

`average <- rnorm(1000) %>%` 

  `enframe()`

`average %>%`

  `ggplot(aes(x = value)) +`

  `geom_histogram()`

` ``` `

_end code chunk_

To run the code chunks: set the cursor between the start ` ```{r} ` and the end 
` ``` ` of the chunk and press Ctrl, Shift, Enter, or choose "Run" in the top-right lint.

## To start a new script 
you can look under File -> New File -> RMarkdown or R Script. You will see a lot of other options as well. In this course we will not address these.

## Getting started

## R & RStudio
Throughout this course you will be working with the open source programs R and RStudio. RStudio is an integrated development environment (IDE) that allows you to use R in an interactive (and graphical) way, write your scripts/code and manage your workspace. R and Rstudio are already pre-installed on the course laptops, but for home use can be downloaded for free from http://www.r-project.org and http://www.rstudio.com.

## Github account and setting github credentials in RStudio

 - Create an account for www.github.com
 - Install the package `{gitr}` (M. Teunis, 2019) using the following code:
 
```{r, eval = FALSE} 
devtools::install_github("uasghogeschoolutrecht/gitr") 
```

## Getting the materials for the course (R scripts):
```{r, include=FALSE}
#library(tidyverse)
urls <- readr::read_csv(
        here::here(
                "urls"
        )
)

urls[3,1] %>% as.character() 
```

 1. Go the following url: `r urls[3,1]`
 2. Copy the url address to the clipboard
 3. Start a new RStudio Project `File` -> `New Project`
 4. Select `Version Control` -> `Git` 
 5. Paste the url in the `Repository URL` field
 6. click `Create Project` and let the clone finish
 
**YOU NOW HAVE THE FILES TO START THE FIRST LESSONS** 

The accompanying Bookdown Webpages to this course can be found at:

`r urls[2,1]`

## Cloud server for RStudio
During the course, we will be using a preinstalled version of the RStudio-IDE (integrated development environment). This IDE version runs on a remote UBUNTU 18.04 Linux server and has most of the packages that are needed to follow the course already installed. You can login at the communicated server address and with the credentials supplied at the beginning of the course. The advantage of cloud servers is that you do not need to install the software on their own laptops. 

The address for the server is: `r urls[4,1]`

To see what the current working directory (wd) is, you can execute the command

`getwd()` 

in the Console window. Try it now!

If all went well the latter half of the working directory will state "/exploratory_data"

## Relative paths
The `getwd()` command gets information on the current wd. Some examples taken from the internet or classical users of R will tell you to use `setwd()` to change the working directory. However, please ignore this advice and always work from an RStudio Project in which you can use relative path. The `{here}` package can help you write clear code to achiev this. See for more details: [`{here}`](https://github.com/jennybc/here_here)

## Creating a new R script
For most analyses/projects, it is best to start by creating an R script file that will contain all the steps taken to get to a certain result. Initially this might be more like a scrapbook to note down what exactly you did. 

Later you can then separate this more cleanly in a workflow (steps taken and/or functions executed) and separate files for function definitions, or even package definitions. 

During this course, we will create special R scripts called RMarkdown (`.Rmd`) for every exercise that contains the steps of that particular exercise. RMarkdown helps you to create a workflow and document your steps. 

Common R scripts contains all functions that can to be reused between different exercises (for instance a file named "functions.R"). 

## Setting up package repositories
Default RStudio has only enabled the CRAN repository, but there are other useful repositories that contain nice R packages. 

```{r, eval = FALSE, echo=TRUE, results='asis'}
setRepositories(graphics=TRUE)
```

## Installing a package from a repository
As an example package (we will be using this later on), we are going to install the "beanplot" package which allows you to easily create so called beanplots. Details of what you can do with them and how to use them, will be provided later during the course. To install the package, go to the Packages window and click on the "install" icon. This will show a popup window where you can type in the name of the package. Type in beanplot and press the "Ok" button. This will download and install the beanplot package.

## Do everything programmatically
The alternative and BETTER WAY is to programmatically install packages:

```{r, eval=FALSE, echo=TRUE}
install.packages("beanplot")
```

To load this library into your workspace, type in library("beanplot") in the Console (or even better, in the R script editor window and then press the "Run" button which will execute the code on that particular line in the R script / or choose `Cntrl` and `Enter` simultaneously).
You can now use the functions defined within this package, for instance to plot a beanplot of (not so useful) random values

```{r, echo=TRUE, fig.show='asis', results='asis'}
library(beanplot)
beanplot(runif(100))
``` 

## Getting help
Knowing how to get help, might be one of the most important skills when learning a new programming language. In R, there are various ways to get help. 

A good start, is the overview documentation provided within R. You can get to this by executing help.start() in the Console window. This will show the help documentation in the Help window within Rstudio. There are a number of good manuals provided within this documentation, particularly useful are the "An Introduction to R" and "The R language definition" manuals. Have a quick look at both.

Besides basic documentation, you also can get package specific documentation. The overview of the different packages and corresponding documentation can be found by clicking on the "Packages" link (You can also get here by going to the Packages window directly). How many packages are user specific?
Lookup the documentation for the function sqrt(), what does it do?
Run the examples for the function sqrt(). What does this do?

## A good resource to start
Free material on learning R is very broadly available on the web. You can start your adventuRe at:

http://r4ds.had.co.nz/index.html

the author Hadley Wickham is an excellent data scientist, speaker and developer at RStudio and the brain behind the "tidyverse". Garret Grolemund is an excellent speaker and the lead educator at RStudio. 

