---
title: "Spare exercises"
author: "Marc Teunis"
date: "12-9-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This dataset is also on risk factors, but now for developing Breast cancer. It was doenloaded from here: https://www.bcsc-research.org/download_file/view/189/344

“Data collection and sharing was supported by the National Cancer Institute (P01CA154292; U54CA163303), the Patient-Centered Outcomes Research Institute (PCS-1504-30370), and the Agency for Health Research and Quality (R01 HS018366-01A1). We thank the participating women, mammography facilities, and radiologists for the data they have provided for this study. You can learn more about the BCSC at: http://www.bcsc-research.org/."


A) Unzip the data, load into R as a data frame (the data is available as "./data/bcsc_risk_factors_summarized1_corrected.zip")

B) inspect the data, the variables, change the names of the variables to safe names if necessary.

C) Check the variable types and change if necessary.

D) Can you create a number of visualization from which you determine a strong riskfactor for the development of Breast cancer from this data?

```{r}
unzip(
  here::here("data",
       "bcsc_risk_factors_summarized1_corrected.zip"),
  exdir = "data", "bcsc_risk_factors_summarized1.csv")

data_bc_risk <- read_csv(
  here::here(
    "data",
    "bcsc_risk_factors_summarized1.csv")
)

data_bc_risk %>%
  is.na() %>%
  sum()

data_bc_risk %>% names

map(
  .x = data_bc_risk,
  .f = typeof
  )


data_bc_risk %>%
  sample_n(100) %>%
  ggplot(aes(x = bmi_group, y = BIRADS_breast_density)) +
  geom_point()
  
```
